---
title: "Preparing CACR Avian Surverys"
author: "WLL"
date: "2026-07-17"
output: html_document
---

# GitHub Setup

I am using GitHub as a form of version control, so I will need to connect this document to the repository on GitHub. While the version set up was already created in tandem to this document, RStudio will often require verification before you can push any committed changes to your code.

**This section is only needed for those modifying code**

```{r Git Setup, eval=FALSE, include = FALSE, message = FALSE, warning = FALSE}

# We will need the usethis package so that we can connect our GitHub

install.packages("usethis")
library(usethis)

# Using the usethis package we can tell R which user will be pushing new versions to the associated repository

use_git_config(user.name = "BLLoftin96", user.email = "brittanieloftin@u.boisestate.edu")

# This will show you who is logged in and will give you options following options: 1) Abort update with error, and keep the existing credentials, 2) Replace these credentials, and 3) See the password/token. If this is a fresh session, I would recommend the replace option and re-enter your current password/token to ensure there is no connection error. 

gitcreds::gitcreds_set()

# This will verify any changes

gitcreds::gitcreds_get()

```


# Install Packages

```{r Install Packages, eval = FALSE}
pkg <- c("dplyr", "tidyr", "tidyverse", "readr", "stringr", "knitr", "rmarkdown", "bibtex", "knitcitations", "bookdown", "sf")
print("Check if packages are installed")
new.pkg <- pkg[!(pkg %in% installed.packages())]
if(length(new.pkg) > 0){
  print(paste("Install missing package(s):", new.pkg, sep=' '))
  install.packages(new.pkg, dependencies = TRUE)
}else{
  print("All packages are already installed!")
}
print("Load packages and return status")
sapply(pkg, require, character.only = TRUE)
```

# Load Libraries

```{r Load Libraries}
library(knitr)
library(rmarkdown)
library(dplyr)
library(tidyr)
library(bibtex)
library(knitcitations)
library(bookdown)
library(tidyverse)
library(readr)
library(stringr)
library(sf)


# Generate BibTex citation file for all R packages used to
# produce report
knitr::write_bib(.packages(), file = "packages.bib")
```

# Setup 

```{r Setup }
#### Setup ####
# Define the project, input data, and cleaned output directories

project_dir <- "R:/CACR_HSM_LiDAR/CACR_Presence.Absence"
data_dir <- file.path(project_dir, "HSM_Data")
output_dir <- file.path(project_dir, "Cleaned_Data")

dir.create(output_dir, showWarnings = FALSE)

#### Create Output Folders ####
dir.create(file.path(output_dir, "Presence_Data"),
           showWarnings = FALSE)

dir.create(file.path(output_dir, "Summaries"),
           showWarnings = FALSE)
```

# Background

The Cassia Crossbill (*Loxia sinesciuris*) is Idaho's only endemic bird and is currently listed as a species of conservation concern. Cassia Crossbills are a unique species that are only found within the high elevation lodgepole pine forests of two mountain ranges of southern Idaho: South Hills and Albion Mountains (Benkman et al. 2009, Behl and Benkman 2018). Like most crossbills, the Cassia Crossbill has overlapping mandible that creating the crossed bill (hence the name) that are specialized to opening a specific type of conifer species or variety (Newton 1972, Benkman 1978a, Benkman 1978b, Benkman 1988a, Benkman 1988b, Benkman 1989, Benkman 1990, Benkman 1992, Benkman 1993, Benkman and Lindholm 1991). In the case of the Cassia Crossbill, they are specialized to feed on the seeds found within the serotinous cones of Rocky Mountain lodgepole pine (*Pinus contorta latifolia*) and are unable to survive competition with other crossbills for other seeds elsewhere (Benkman et al. 2001, Benkman et al. 2012, Benkman 2016). The dependency on a single resource makes the survival and reproduction of crossbills depend on the availability of that resource and they are thus extremely vulnerable to habitat loss and alteration (Newton 1972, Benkman 1987a, Benkman 1988a, Benkman 1990, Benkman 1992). Despite lodgepole pine existing in other areas, the Cassia Crossbill was able to evolve in this region due to the two mountain ranges completely lacking red squirrels (*Tamiasciurus hudsonicus*) which typically preemptively out compete crossbills for cones (Benkman 1999, Smith and Benkman 2007).

With Cassia Crossbills being permanent residents confined to the highest elevations of these two mountain ranges, they are especially susceptible to the effects of anthropogenic climate change and have a high risk of declining population trends and extinction (Parmesan 2006). The combination of being a permanent resident in a restricted range and the dependence on serotinous cones of Rocky Mountain lodgepole pine make the Cassia Crossbill even more vulnerable to habitat loss or alteration compared to more widespread montane species (Benkman 1993). These alterations and habitat loss are predicted to occur under anthropogenic climate change as regional forecasts indicate dramatic shifts in plant distributions with a predicted 50% or greater loss in the distribution of lodgepole pine, including its extirpation from the Cassia Crossbill’s range by 2080 (Thompson et al. 1998, Coops and Waring 2011). Additionally, increasing summer temperatures have already been recorded as having a negative impact on Cassia Crossbills as it causes their primary food source, serotinous lodgepole cones, to prematurely open allowing other species to access this once exclusive food source (Benkman et al. 2001, Benkman et al. 2012, Santisteban et al. 2012, Benkman 2016). Cone crops have also been affected by the increase in fires in this area either through being directly destroyed or by causing cones to open and allowing other species to access the seeds. Fires will only become more problematic in the future as they are predicted to increase in frequency and intensity due to multiple interacting factors of climate change (Brice et al. 2020, Rogers et al. 2020).

The Cassia Crossbill Working Group (CCWG) which formed in early 2021 from a partnership of 22 stakeholders including members from Boise State University Human-Environment Systems, Idaho Department of Fish and Game (IDFG), Intermountain Bird Observatory (IBO), the Sawtooth National Forest, the Shoshone-Bannock Tribes, and U.S Fish and Wildlife Service. The goal of the CCWG is to work collaboratively on conservation planning, habitat restoration, and long-term monitoring of the Cassia Crossbill. As part of IBO's agreement with the forest service and IDFG, annual surveys have been conducted since 2021 to estimate the population estimates using sites that had been used by Dr. Craig Benkman in his previous studies. However, one of the most needed products identified by managers part of the CCWG was a habitat suitability model (HSM) of the Cassia Crossbill.

# Sampling Design

To create an accurate HSM, surveys needed to be conducted across the ecological gradients of environments available to Cassia Crossbills which wasn't captured in the previous sampling scheme as areas were only sampling Lodgepole Pine designated areas by the Behl Model (Behl and Benkman 2018) or areas of "highest quality habitat" ascribed by Dr. Benkman. It was decided in 2023 that we needed to expand and refine our sampling scheme to get a more accurate model. Therefore, Ryan Healey from IBO set out to create a new lodgepole pine model to improve areas that were designated as lodgepole pine areas. Once the model was created we used GRTS to get a random stratified sample of these points and selected approximately 300 lodgepole pine sites for 2023 and followed a similar method to select approximately 50 non-lodgepole sites. In 2024, Ryan Healey used the data from 2023 to update the lodgepole model and, using the updated version, new sampling sites were selected in 2024 following the same protocol as 2023 for site selection.

At each site, surveyors would conduct a 10-minute standardized survey broken into two 5-minute intervals noting any crossbills and woodpeckers seen or heard. If a bird was seen/heard, the surveyor would record the initial distance, bearing, interval, flock size, and whether the bird was perched or a flyover. We recorded whether the bird was perched or flying over because we want to ensure crossbills are actually using the habitat for the model to be accurate and crossbills are known to fly long distances to get to resources. Additionally, at each point vegetation surveys were conducted recording Integrated Monitoring in Bird Conservation Regions (IMBCR) standardized measurements within 50m of the point. Vegetation surveys included measurements of both canopy and understory to help gain an understanding of the habitat most suitable for CACR. **Currently, we have LiDAR data for the study areas from the Idaho LiDAR Consortium so we want to try to use this data as a way to infer forest metrics and determine if any forest structures influence the suitability of landscape for Cassia Crossbills.**

# Goal

Overall, we want to create two habitat suitability models (difference described below) that incorporate Cassia Crossbill presence/absence data collected during our annual surveys, forest metrics from the LiDAR data, and existing elevation metrics.

1.  Conservative Model: Presence of Cassia Crossbills is only indicated by the presence of a perched bird being directly observed by a surveyor.

2.  Liberal Model: Presence of Cassia Crossbills is indicated by either a perched bird observed by a surveyor AND/OR the siting of crossbill-opened serotinous lodgepole pine cones

To complete this goal, we will need to use the 2023 and 2024 Cassia Crossbill surveys to build our model then use the 2025 summer and fall data to validate and update the model. We should, first, clean and sort the 2023 and 2024 data to create our two types of data frames (Conservative_Presence vs. Liberal_Presence) that will build the two models. Then, we will want to take the LiDAR data - which should only contain the tiles for our study area (*to see how this was made see the Study_Area_LiDAR_Setup*) - and validate the data to ensure it looks correct and makes sense. Once this is done, we will stitch the tiles together. Then, we will want to decide on our Forest metrics ensuring that they make biological sense and make rasters containing the metrics. We will want to gather any other spatial data or rasters (e.g., elevation, slope, aspect) and generate a raster stack for our model. Finally, we can create our two habitat suitability models and test how accurate it is.

# Data Cleaning

## 2023

```{r Clean and Classify 2023 Survey Data}
# Clean 2023 bird observations, create strict and liberal presence datasets, and evaluate how a 50 m cutoff would affect perched CACR detections
# We cannot be certain that techs entered both species type and detection of CACR (yes or no) correctly every time - I expect some human error.

#### Load Data ####
counts_2023 <- read_csv(
  file.path(data_dir, "2023", "CACRCounts2023.csv"),
  show_col_types = FALSE
) |>
  transmute(
    Site = `Survey Point Number`,
    Date,
    Species = str_trim(str_to_upper(Species)),
    Distance = `Initial Distance`,
    Detection_Type = str_trim(str_to_upper(`Perched or Flyover`)),
    x,
    y
  ) |>
  mutate(
    Strict_Detection = Species == "CACR" &
      Detection_Type == "PERCHED"
  )

veg_raw_2023 <- read_csv(
  file.path(data_dir, "2023", "veg.csv"),
  show_col_types = FALSE,
  name_repair = "unique"
)

#### Strict 2023 Presence Data ####

strict_2023 <- counts_2023 |>
  group_by(Site) |>
  summarise(
    x = first(x[!is.na(x)], default = NA_real_),
    y = first(y[!is.na(y)], default = NA_real_),

    Strict_Presence = as.integer(
      any(Strict_Detection, na.rm = TRUE)
    ),

    Min_Perched_Distance = if (
      any(Strict_Detection & !is.na(Distance), na.rm = TRUE)
    ) {
      min(
        Distance[Strict_Detection & !is.na(Distance)],
        na.rm = TRUE
      )
    } else {
      NA_real_
    },

    Strict_Presence_50m = case_when(
      !any(Strict_Detection, na.rm = TRUE) ~ 0L,

      any(
        Strict_Detection &
          !is.na(Distance) &
          Distance <= 50,
        na.rm = TRUE
      ) ~ 1L,

      any(
        Strict_Detection &
          !is.na(Distance),
        na.rm = TRUE
      ) ~ 0L,

      TRUE ~ NA_integer_
    ),

    .groups = "drop"
  ) |>
  mutate(
    Changed_By_50m = Strict_Presence == 1 &
      Strict_Presence_50m == 0
  )


#### 2023 Liberal Presence Data ####
# Identify crossbill-foraged cones on any sampled branch and combine that evidence with perched CACR detections

cone_foraging_columns_2023 <- names(veg_raw_2023)[
  str_detect(
    names(veg_raw_2023),
    regex(
      "^Branch [1-5] Percent Cones Foraged",
      ignore_case = TRUE
    )
  )
]

cat(
  "Crossbill-foraging branch columns found:",
  length(cone_foraging_columns_2023),
  "\n"
)

cone_evidence_2023 <- veg_raw_2023 |>
  transmute(
    Site = `Survey Point Number`,

    Cone_Evidence = if_any(
      all_of(cone_foraging_columns_2023),
      ~ !is.na(.x) & .x > 0
    )
  ) |>
  group_by(Site) |>
  summarise(
    Cone_Evidence = any(Cone_Evidence, na.rm = TRUE),
    .groups = "drop"
  )

liberal_2023 <- strict_2023 |>
  left_join(
    cone_evidence_2023,
    by = "Site"
  ) |>
  mutate(
    Cone_Evidence = replace_na(
      Cone_Evidence,
      FALSE
    ),

    Liberal_Presence = as.integer(
      Strict_Presence == 1 |
        Cone_Evidence
    ),

    Presence_Evidence = case_when(
      Strict_Presence == 1 & Cone_Evidence ~
        "Perched CACR and cone evidence",

      Strict_Presence == 1 ~
        "Perched CACR only",

      Cone_Evidence ~
        "Cone evidence only",

      TRUE ~
        "No presence evidence"
    )
  )


#### Summarize 2023 Presence Classifications ####
# Report strict and liberal presence counts and the type of evidence supporting each liberal presence classification

cat("Total survey sites:", nrow(liberal_2023), "\n")

cat(
  "Strict presences:",
  sum(liberal_2023$Strict_Presence, na.rm = TRUE),
  "\n"
)

cat(
  "Strict absences:",
  sum(liberal_2023$Strict_Presence == 0, na.rm = TRUE),
  "\n"
)

cat(
  "Liberal presences:",
  sum(liberal_2023$Liberal_Presence, na.rm = TRUE),
  "\n"
)

cat(
  "Liberal absences:",
  sum(liberal_2023$Liberal_Presence == 0, na.rm = TRUE),
  "\n"
)

print(
  liberal_2023 |>
    count(
      Presence_Evidence,
      name = "Sites"
    )
)


#### Evaluate 2023 50 m Detection Cutoff ####
# Summarize individual perched CACR detections and site classifications with and without a 50 m cutoff

perched_detection_50m_summary_2023 <- counts_2023 |>
  filter(Strict_Detection) |>
  mutate(
    Distance_Category = case_when(
      is.na(Distance) ~ "Missing distance",
      Distance <= 50 ~ "Within 50 m",
      Distance > 50 ~ "Beyond 50 m"
    )
  ) |>
  count(
    Distance_Category,
    name = "Detections"
  ) |>
  mutate(
    Survey_Period = "2023",
    .before = 1
  )

strict_site_50m_summary_2023 <- strict_2023 |>
  summarise(
    Survey_Period = "2023",

    Strict_Presence_No_Cutoff = sum(
      Strict_Presence,
      na.rm = TRUE
    ),

    Strict_Presence_Within_50m = sum(
      Strict_Presence_50m == 1,
      na.rm = TRUE
    ),

    Sites_Removed_By_50m = sum(
      Changed_By_50m,
      na.rm = TRUE
    ),

    Sites_With_Missing_Perched_Distance = sum(
      Strict_Presence == 1 &
        is.na(Strict_Presence_50m),
      na.rm = TRUE
    )
  )

print(perched_detection_50m_summary_2023)
print(strict_site_50m_summary_2023)

#### Save 2023 Outputs ####
# Write cleaned 2023 datasets and summaries to CSV files

write_csv(
  strict_2023,
  file.path(
    output_dir,
    "Presence_Data",
    "Strict_Presence_2023.csv"
  )
)

write_csv(
  liberal_2023,
  file.path(
    output_dir,
    "Presence_Data",
    "Liberal_Presence_2023.csv"
  )
)

write_csv(
  perched_detection_50m_summary_2023,
  file.path(
    output_dir,
    "Summaries",
    "Detection_50m_Summary_2023.csv"
  )
)

write_csv(
  strict_site_50m_summary_2023,
  file.path(
    output_dir,
    "Summaries",
    "Site_50m_Summary_2023.csv"
  )
)
```

Applying the 50m cutoff would remove 18 sites or 34.6% of the strict data.

## 2024

```{r Clean and Classify 2024 Survey Data}
# Clean 2024 bird observations, create strict and liberal presence datasets, and evaluate how a 50 m cutoff would affect perched CACR detections
# Presence is based on the recorded species and whether the CACR detection was perched or a flyover.

#### Load Data ####

counts_2024 <- read_csv(
  file.path(data_dir, "2024", "AllcountData.csv"),
  show_col_types = FALSE
) |>
  transmute(
    Site = Survey.Point.ID,
    Species = str_trim(str_to_upper(Species)),
    Distance = Initial.Distance,
    Detection_Type = str_trim(str_to_upper(Perched.or.Flyover)),
    x,
    y
  ) |>
  mutate(
    Strict_Detection = Species == "CACR" &
      Detection_Type == "PERCHED"
  )

veg_raw_2024 <- read_csv(
  file.path(data_dir, "2024", "CACR2024_LodgepoleSurveys.csv"),
  show_col_types = FALSE,
  name_repair = "unique"
)


#### Strict 2024 Presence Data ####

strict_2024 <- counts_2024 |>
  group_by(Site) |>
  summarise(
    x = first(x[!is.na(x)], default = NA_real_),
    y = first(y[!is.na(y)], default = NA_real_),

    Strict_Presence = as.integer(
      any(Strict_Detection, na.rm = TRUE)
    ),

    Min_Perched_Distance = if (
      any(Strict_Detection & !is.na(Distance), na.rm = TRUE)
    ) {
      min(
        Distance[Strict_Detection & !is.na(Distance)],
        na.rm = TRUE
      )
    } else {
      NA_real_
    },

    Strict_Presence_50m = case_when(
      !any(Strict_Detection, na.rm = TRUE) ~ 0L,

      any(
        Strict_Detection &
          !is.na(Distance) &
          Distance <= 50,
        na.rm = TRUE
      ) ~ 1L,

      any(
        Strict_Detection &
          !is.na(Distance),
        na.rm = TRUE
      ) ~ 0L,

      TRUE ~ NA_integer_
    ),

    .groups = "drop"
  ) |>
  mutate(
    Changed_By_50m = Strict_Presence == 1 &
      Strict_Presence_50m == 0
  )


#### 2024 Liberal Presence Data ####
# Identify crossbill-foraged cones on any sampled branch and combine that evidence with perched CACR detections

cone_foraging_columns_2024 <- names(veg_raw_2024)[
  str_detect(
    names(veg_raw_2024),
    regex(
      "^% (sampled )?foraged cones|^% foraged cones \\| branch",
      ignore_case = TRUE
    )
  )
]

cat(
  "Crossbill-foraging branch columns found:",
  length(cone_foraging_columns_2024),
  "\n"
)

print(cone_foraging_columns_2024)

cone_evidence_2024 <- veg_raw_2024 |>
  transmute(
    Site = `Survey Point ID`,

    Cone_Evidence = if_any(
      all_of(cone_foraging_columns_2024),
      ~ !is.na(.x) & .x > 0
    )
  ) |>
  group_by(Site) |>
  summarise(
    Cone_Evidence = any(Cone_Evidence, na.rm = TRUE),
    .groups = "drop"
  )

liberal_2024 <- strict_2024 |>
  left_join(
    cone_evidence_2024,
    by = "Site"
  ) |>
  mutate(
    Cone_Evidence = replace_na(
      Cone_Evidence,
      FALSE
    ),

    Liberal_Presence = as.integer(
      Strict_Presence == 1 |
        Cone_Evidence
    ),

    Presence_Evidence = case_when(
      Strict_Presence == 1 & Cone_Evidence ~
        "Perched CACR and cone evidence",

      Strict_Presence == 1 ~
        "Perched CACR only",

      Cone_Evidence ~
        "Cone evidence only",

      TRUE ~
        "No presence evidence"
    )
  )


#### Summarize 2024 Presence Classifications ####
# Report strict and liberal presence counts and the type of evidence supporting each liberal presence classification

cat("Total survey sites:", nrow(liberal_2024), "\n")

cat(
  "Strict presences:",
  sum(liberal_2024$Strict_Presence, na.rm = TRUE),
  "\n"
)

cat(
  "Strict absences:",
  sum(liberal_2024$Strict_Presence == 0, na.rm = TRUE),
  "\n"
)

cat(
  "Liberal presences:",
  sum(liberal_2024$Liberal_Presence, na.rm = TRUE),
  "\n"
)

cat(
  "Liberal absences:",
  sum(liberal_2024$Liberal_Presence == 0, na.rm = TRUE),
  "\n"
)

print(
  liberal_2024 |>
    count(
      Presence_Evidence,
      name = "Sites"
    )
)


#### Evaluate 2024 50 m Detection Cutoff ####
# Summarize individual perched CACR detections and site classifications with and without a 50 m cutoff

perched_detection_50m_summary_2024 <- counts_2024 |>
  filter(Strict_Detection) |>
  mutate(
    Distance_Category = case_when(
      is.na(Distance) ~ "Missing distance",
      Distance <= 50 ~ "Within 50 m",
      Distance > 50 ~ "Beyond 50 m"
    )
  ) |>
  count(
    Distance_Category,
    name = "Detections"
  ) |>
  mutate(
    Survey_Period = "2024",
    .before = 1
  )

strict_site_50m_summary_2024 <- strict_2024 |>
  summarise(
    Survey_Period = "2024",

    Strict_Presence_No_Cutoff = sum(
      Strict_Presence,
      na.rm = TRUE
    ),

    Strict_Presence_Within_50m = sum(
      Strict_Presence_50m == 1,
      na.rm = TRUE
    ),

    Sites_Removed_By_50m = sum(
      Changed_By_50m,
      na.rm = TRUE
    ),

    Sites_With_Missing_Perched_Distance = sum(
      Strict_Presence == 1 &
        is.na(Strict_Presence_50m),
      na.rm = TRUE
    )
  )

print(perched_detection_50m_summary_2024)
print(strict_site_50m_summary_2024)


#### Save 2024 Outputs ####
# Write cleaned 2024 datasets and summaries to CSV files

write_csv(
  strict_2024,
  file.path(
    output_dir,
    "Presence_Data",
    "Strict_Presence_2024.csv"
  )
)

write_csv(
  liberal_2024,
  file.path(
    output_dir,
    "Presence_Data",
    "Liberal_Presence_2024.csv"
  )
)

write_csv(
  perched_detection_50m_summary_2024,
  file.path(
    output_dir,
    "Summaries",
    "Detection_50m_Summary_2024.csv"
  )
)

write_csv(
  strict_site_50m_summary_2024,
  file.path(
    output_dir,
    "Summaries",
    "Site_50m_Summary_2024.csv"
  )
)
```
Removing detections over 50m would remove ~31% of the presence strict data.


## 2025

### Summer

```{r Clean and Classify Summer 2025 Survey Data}
# I noticed a duplicate site ID for MAC-124 but in different areas, so I will be using a shapefile to validate if this is a typo. I will need to standardize the thermologger shapefile

#### Load Summer 2025 Data and Check MAC-124 ####

survey_summer_2025 <- read_csv(
  file.path(
    data_dir,
    "2025",
    "Summer",
    "2025_CACR_Summer_Survey.csv"
  ),
  show_col_types = FALSE
)

cone_counts_summer_2025 <- read_csv(
  file.path(
    data_dir,
    "2025",
    "Summer",
    "2025_Summer_Cone_Counts.csv"
  ),
  show_col_types = FALSE
)

cone_survey_summer_2025 <- read_csv(
  file.path(
    data_dir,
    "2025",
    "Summer",
    "2025_Summer_Cone_Survey.csv"
  ),
  show_col_types = FALSE
)

logger_cones_summer_2025 <- read_csv(
  file.path(
    data_dir,
    "2025",
    "Summer",
    "2025_Cone_Surveys_Logger_Already_Deployed.csv"
  ),
  show_col_types = FALSE
)


# Replace this filename with the official survey-site shapefile name

site_shapefile <- file.path(
  project_dir,
  "Thermologger_Deployment_Locations_2025.shp"
)

official_sites_raw <- st_read(
  site_shapefile,
  quiet = TRUE
)


# Replace this value only if the official site-ID field has a different name

site_id_field <- "name"

if (!site_id_field %in% names(official_sites_raw)) {
  stop(
    paste0(
      "The field '",
      site_id_field,
      "' was not found in the shapefile.\n",
      "Available fields are:\n",
      paste(names(official_sites_raw), collapse = ", ")
    )
  )
}

if (is.na(st_crs(official_sites_raw))) {
  stop(
    "The official site shapefile does not have a defined coordinate reference system."
  )
}


# Standardize official IDs by removing the trailing "-D" used only in the shapefile

official_sites <- official_sites_raw |>
  mutate(
    Official_Site_Full = str_trim(
      str_to_upper(
        as.character(.data[[site_id_field]])
      )
    ),

    Official_Site = str_remove(
      Official_Site_Full,
      "-D$"
    )
  ) |>
  filter(
    !is.na(Official_Site),
    Official_Site != ""
  )


# Locate the records entered as MAC-124

mac124_records <- survey_summer_2025 |>
  mutate(
    Survey_Row = row_number(),

    Recorded_Site = str_trim(
      str_to_upper(
        as.character(`Survey Point Number`)
      )
    )
  ) |>
  filter(
    Recorded_Site == "MAC-124"
  )

cat(
  "Records entered as MAC-124:",
  nrow(mac124_records),
  "\n"
)

if (nrow(mac124_records) != 2) {
  warning(
    paste0(
      "Expected two records entered as MAC-124 but found ",
      nrow(mac124_records),
      ". The location check will still continue."
    )
  )
}


# Convert the survey coordinates to spatial points and match each MAC-124 record to the nearest official survey location

mac124_points <- mac124_records |>
  filter(
    !is.na(x),
    !is.na(y)
  ) |>
  st_as_sf(
    coords = c("x", "y"),
    crs = 4326,
    remove = FALSE
  ) |>
  st_transform(
    st_crs(official_sites)
  )

nearest_official_row <- st_nearest_feature(
  mac124_points,
  official_sites
)

mac124_location_check <- mac124_points |>
  mutate(
    Nearest_Official_Site_Full =
      official_sites$Official_Site_Full[
        nearest_official_row
      ],

    Nearest_Official_Site =
      official_sites$Official_Site[
        nearest_official_row
      ],

    Distance_to_Official_Site_m = as.numeric(
      st_distance(
        mac124_points,
        official_sites[
          nearest_official_row,
        ],
        by_element = TRUE
      )
    ),

    Site_ID_Matches_Location =
      Recorded_Site == Nearest_Official_Site
  ) |>
  st_drop_geometry() |>
  select(
    Survey_Row,
    `Date and Time`,
    Recorded_Site,
    Nearest_Official_Site,
    Nearest_Official_Site_Full,
    Distance_to_Official_Site_m,
    Site_ID_Matches_Location,
    x,
    y,
    everything()
  )

print(
  mac124_location_check |>
    select(
      Survey_Row,
      `Date and Time`,
      Recorded_Site,
      Nearest_Official_Site,
      Nearest_Official_Site_Full,
      Distance_to_Official_Site_m,
      Site_ID_Matches_Location,
      x,
      y
    )
)


# Automatically correct a MAC-124 record only when:
# 1. its nearest official site has a different ID, and
# 2. the recorded coordinates are within 250 m of that official point.
#
# Records farther than 250 m are retained unchanged and flagged for manual review.

max_match_distance_m <- 250

mac124_corrections <- mac124_location_check |>
  transmute(
    Survey_Row,

    Corrected_Site = case_when(
      Nearest_Official_Site != Recorded_Site &
        Distance_to_Official_Site_m <=
          max_match_distance_m ~
        Nearest_Official_Site,

      TRUE ~
        Recorded_Site
    ),

    Site_ID_Corrected =
      Corrected_Site != Recorded_Site,

    Requires_Manual_Review =
      Distance_to_Official_Site_m >
        max_match_distance_m
  )


survey_summer_2025 <- survey_summer_2025 |>
  mutate(
    Survey_Row = row_number(),

    Site_Originally_Recorded = str_trim(
      str_to_upper(
        as.character(`Survey Point Number`)
      )
    )
  ) |>
  left_join(
    mac124_corrections,
    by = "Survey_Row"
  ) |>
  mutate(
    Corrected_Site = coalesce(
      Corrected_Site,
      Site_Originally_Recorded
    ),

    Site_ID_Corrected = replace_na(
      Site_ID_Corrected,
      FALSE
    ),

    Requires_Manual_Review = replace_na(
      Requires_Manual_Review,
      FALSE
    ),

    `Survey Point Number` = Corrected_Site
  ) |>
  select(
    -Survey_Row,
    -Corrected_Site
  )


# Show any changes made to the original survey records

mac124_correction_summary <- survey_summer_2025 |>
  filter(
    Site_Originally_Recorded == "MAC-124"
  ) |>
  select(
    `Date and Time`,
    Site_Originally_Recorded,
    Corrected_Site = `Survey Point Number`,
    Site_ID_Corrected,
    Requires_Manual_Review,
    x,
    y
  )

print(mac124_correction_summary)

cat(
  "MAC-124 records corrected:",
  sum(
    mac124_correction_summary$Site_ID_Corrected,
    na.rm = TRUE
  ),
  "\n"
)

#### Create Strict Summer 2025 Presence Data ####
# Strict presence is based on whether Cassia Crossbills were directly detected.

counts_summer_2025 <- survey_summer_2025 |>
  transmute(
    Site = str_trim(
      str_to_upper(
        as.character(`Survey Point Number`)
      )
    ),

    Survey_Date = `Date and Time`,

    CACR_Detected_Response = str_trim(
      str_to_upper(
        as.character(
          `Any Cassia Crossbills Detected`
        )
      )
    ),

    x,
    y
  )


# Check that every survey has a usable CACR detection response

missing_detection_responses_summer_2025 <-
  counts_summer_2025 |>
  filter(
    is.na(CACR_Detected_Response) |
      !CACR_Detected_Response %in%
        c(
          "YES",
          "NO"
        )
  )

if (nrow(missing_detection_responses_summer_2025) > 0) {
  print(
    missing_detection_responses_summer_2025
  )

  stop(
    paste0(
      nrow(missing_detection_responses_summer_2025),
      " Summer 2025 survey record(s) have a missing or unexpected ",
      "CACR detection response. Review the printed records before continuing."
    )
  )
}


counts_summer_2025 <- counts_summer_2025 |>
  mutate(
    CACR_Detected =
      CACR_Detected_Response == "YES"
  )


strict_summer_2025 <- counts_summer_2025 |>
  group_by(
    Site
  ) |>
  summarise(
    x = first(
      x[
        !is.na(x)
      ],
      default = NA_real_
    ),

    y = first(
      y[
        !is.na(y)
      ],
      default = NA_real_
    ),

    Survey_Visits = n(),

    Strict_Presence = as.integer(
      any(
        CACR_Detected
      )
    ),

    .groups = "drop"
  ) |>
  arrange(
    Site
  )


#### Create Minimal Summer 2025 Cone-Evidence Data ####
# Liberal presence requires only whether any crossbill-opened cones were found.
# Crossbill-opened cones may occur in either the canopy or at the tree base.


# Create a lookup connecting additional-tree cone records to survey sites

cone_survey_site_lookup_summer_2025 <-
  cone_survey_summer_2025 |>
  transmute(
    ParentGlobalID =
      GlobalID,

    Site = str_remove(
      str_trim(
        str_to_upper(
          as.character(
            `Survey Point ID`
          )
        )
      ),
      "-D$"
    )
  )


# Join additional-tree cone counts to their survey sites

additional_cone_records_summer_2025 <-
  cone_counts_summer_2025 |>
  left_join(
    cone_survey_site_lookup_summer_2025,
    by = "ParentGlobalID"
  )


# Every additional-tree record should match a parent cone survey

unmatched_additional_cones_summer_2025 <-
  additional_cone_records_summer_2025 |>
  filter(
    is.na(Site) |
      Site == ""
  )

if (nrow(unmatched_additional_cones_summer_2025) > 0) {
  print(
    unmatched_additional_cones_summer_2025
  )

  stop(
    paste0(
      nrow(unmatched_additional_cones_summer_2025),
      " additional-tree cone record(s) did not match a parent cone survey. ",
      "Review ParentGlobalID and GlobalID before continuing."
    )
  )
}


# Retain only the information needed to identify cone evidence

additional_cone_evidence_summer_2025 <-
  additional_cone_records_summer_2025 |>
  transmute(
    Site,

    Canopy_Crossbill_Opened_Cones =
      as.numeric(
        `Canopy Crossbill Opened Cones`
      ),

    Base_Crossbill_Opened_Cones =
      as.numeric(
        `Base Crossbill Opened Cones`
      ),

    Cone_Evidence =
      coalesce(
        Canopy_Crossbill_Opened_Cones,
        0
      ) > 0 |
      coalesce(
        Base_Crossbill_Opened_Cones,
        0
      ) > 0
  )


# Prepare cone evidence from thermologger trees

logger_cone_evidence_summer_2025 <-
  logger_cones_summer_2025 |>
  transmute(
    Site = str_remove(
      str_trim(
        str_to_upper(
          as.character(
            `Survey Point ID`
          )
        )
      ),
      "-D$"
    ),

    Canopy_Crossbill_Opened_Cones =
      as.numeric(
        `Canopy Crossbill Forage Count`
      ),

    Base_Crossbill_Opened_Cones =
      as.numeric(
        `Base Crossbill Forage Count`
      ),

    Cone_Evidence =
      coalesce(
        Canopy_Crossbill_Opened_Cones,
        0
      ) > 0 |
      coalesce(
        Base_Crossbill_Opened_Cones,
        0
      ) > 0
  )


# Combine all trees and summarize cone evidence by survey site

cone_evidence_summer_2025 <-
  bind_rows(
    logger_cone_evidence_summer_2025,
    additional_cone_evidence_summer_2025
  ) |>
  filter(
    !is.na(Site),
    Site != ""
  ) |>
  group_by(
    Site
  ) |>
  summarise(
    Trees_With_Cone_Counts = n(),

    Trees_With_Cone_Evidence = sum(
      Cone_Evidence
    ),

    Cone_Evidence = any(
      Cone_Evidence
    ),

    .groups = "drop"
  )


#### Check Summer 2025 Cone Coverage ####
# Every bird-survey site should have corresponding cone-count data.
# First run found typos and instances where we had bird surveys with no cone date or vice versa. Specifically, MAC-079 was the label for the cone count but CACR and veg had it labeled as MAC-072 so we will rename it to match. For no cone counts, we had MAC-058 which had an unauthorized camp with activity making it too dangerous to set up a logger, LPO-060 we couldn't find a logger so no cones were surveyed, and MAC-089 had no lodgepole so there were no cones to survey. 

bird_sites_without_cone_counts_summer_2025 <-
  strict_summer_2025 |>
  distinct(
    Site
  ) |>
  anti_join(
    cone_evidence_summer_2025 |>
      distinct(
        Site
      ),
    by = "Site"
  )


cone_sites_without_bird_surveys_summer_2025 <-
  cone_evidence_summer_2025 |>
  distinct(
    Site
  ) |>
  anti_join(
    strict_summer_2025 |>
      distinct(
        Site
      ),
    by = "Site"
  )


cat(
  "Bird-survey sites without cone counts:",
  nrow(
    bird_sites_without_cone_counts_summer_2025
  ),
  "\n"
)

print(
  bird_sites_without_cone_counts_summer_2025
)


cat(
  "Cone-count sites without a bird survey:",
  nrow(
    cone_sites_without_bird_surveys_summer_2025
  ),
  "\n"
)

print(
  cone_sites_without_bird_surveys_summer_2025
)

#### Correct Summer 2025 Cone-Survey Site ID ####
# The cone survey labeled MAC-079 corresponds to the bird and vegetation survey site MAC-072. Correct the cone-survey site ID before summarizing cone evidence.

cone_survey_summer_2025 <- cone_survey_summer_2025 |>
  mutate(
    `Survey Point ID` = case_when(
      str_remove(
        str_trim(
          str_to_upper(
            as.character(`Survey Point ID`)
          )
        ),
        "-D$"
      ) == "MAC-079" ~
        "MAC-072",

      TRUE ~
        as.character(`Survey Point ID`)
    )
  )


logger_cones_summer_2025 <- logger_cones_summer_2025 |>
  mutate(
    `Survey Point ID` = case_when(
      str_remove(
        str_trim(
          str_to_upper(
            as.character(`Survey Point ID`)
          )
        ),
        "-D$"
      ) == "MAC-079" ~
        "MAC-072",

      TRUE ~
        as.character(`Survey Point ID`)
    )
  )


#### Recreate Minimal Summer 2025 Cone-Evidence Data ####
# Match child cone-count records to parent cone surveys using:

# cone_counts_summer_2025$ParentGlobalID = cone_survey_summer_2025$GlobalID

cone_survey_site_lookup_summer_2025 <-
  cone_survey_summer_2025 |>
  transmute(
    ParentGlobalID =
      GlobalID,

    Site = str_remove(
      str_trim(
        str_to_upper(
          as.character(
            `Survey Point ID`
          )
        )
      ),
      "-D$"
    ),

    Cone_x = x,
    Cone_y = y
  )


additional_cone_records_summer_2025 <-
  cone_counts_summer_2025 |>
  left_join(
    cone_survey_site_lookup_summer_2025,
    by = "ParentGlobalID"
  )


unmatched_additional_cones_summer_2025 <-
  additional_cone_records_summer_2025 |>
  filter(
    is.na(Site) |
      Site == ""
  )


if (nrow(unmatched_additional_cones_summer_2025) > 0) {
  print(
    unmatched_additional_cones_summer_2025
  )

  stop(
    paste0(
      nrow(unmatched_additional_cones_summer_2025),
      " additional-tree cone record(s) did not match a parent cone survey. ",
      "Review ParentGlobalID and GlobalID before continuing."
    )
  )
}


additional_cone_evidence_summer_2025 <-
  additional_cone_records_summer_2025 |>
  transmute(
    Site,

    Canopy_Crossbill_Opened_Cones =
      as.numeric(
        `Canopy Crossbill Opened Cones`
      ),

    Base_Crossbill_Opened_Cones =
      as.numeric(
        `Base Crossbill Opened Cones`
      ),

    Cone_Evidence =
      coalesce(
        Canopy_Crossbill_Opened_Cones,
        0
      ) > 0 |
      coalesce(
        Base_Crossbill_Opened_Cones,
        0
      ) > 0
  )


logger_cone_evidence_summer_2025 <-
  logger_cones_summer_2025 |>
  transmute(
    Site = str_remove(
      str_trim(
        str_to_upper(
          as.character(
            `Survey Point ID`
          )
        )
      ),
      "-D$"
    ),

    Canopy_Crossbill_Opened_Cones =
      as.numeric(
        `Canopy Crossbill Forage Count`
      ),

    Base_Crossbill_Opened_Cones =
      as.numeric(
        `Base Crossbill Forage Count`
      ),

    Cone_Evidence =
      coalesce(
        Canopy_Crossbill_Opened_Cones,
        0
      ) > 0 |
      coalesce(
        Base_Crossbill_Opened_Cones,
        0
      ) > 0
  )


cone_evidence_summer_2025 <-
  bind_rows(
    logger_cone_evidence_summer_2025,
    additional_cone_evidence_summer_2025
  ) |>
  filter(
    !is.na(Site),
    Site != ""
  ) |>
  group_by(
    Site
  ) |>
  summarise(
    Trees_With_Cone_Counts = n(),

    Trees_With_Cone_Evidence = sum(
      Cone_Evidence,
      na.rm = TRUE
    ),

    Cone_Evidence = any(
      Cone_Evidence,
      na.rm = TRUE
    ),

    .groups = "drop"
  )


#### Document Summer 2025 Sites Without Cone Surveys ####
# These sites legitimately lack cone-count records for known field reasons.

cone_survey_exceptions_summer_2025 <- tibble(
  Site = c(
    "MAC-058",
    "LPO-060",
    "MAC-089"
  ),

  Cone_Survey_Status = c(
    "Not surveyed",
    "Not surveyed",
    "Not applicable"
  ),

  Cone_Survey_Reason = c(
    paste(
      "Unauthorized camp and associated activity made it",
      "unsafe to establish a thermologger or conduct the cone survey"
    ),

    paste(
      "The previously deployed thermologger could not be located,",
      "so the cone survey was not completed"
    ),

    paste(
      "No lodgepole pine was present at the survey site,",
      "so there were no cones to survey"
    )
  )
)


#### Check Summer 2025 Cone Coverage ####
# Stop only when a bird-survey site lacks cone data and is not one of the documented field exceptions.

bird_sites_without_cone_counts_summer_2025 <-
  strict_summer_2025 |>
  distinct(
    Site
  ) |>
  anti_join(
    cone_evidence_summer_2025 |>
      distinct(
        Site
      ),
    by = "Site"
  ) |>
  left_join(
    cone_survey_exceptions_summer_2025,
    by = "Site"
  )


unexplained_missing_cone_sites_summer_2025 <-
  bird_sites_without_cone_counts_summer_2025 |>
  filter(
    is.na(Cone_Survey_Status)
  )


cone_sites_without_bird_surveys_summer_2025 <-
  cone_evidence_summer_2025 |>
  distinct(
    Site
  ) |>
  anti_join(
    strict_summer_2025 |>
      distinct(
        Site
      ),
    by = "Site"
  )


cat(
  "Bird-survey sites without cone counts:",
  nrow(
    bird_sites_without_cone_counts_summer_2025
  ),
  "\n"
)

print(
  bird_sites_without_cone_counts_summer_2025
)


cat(
  "Unexplained bird-survey sites without cone counts:",
  nrow(
    unexplained_missing_cone_sites_summer_2025
  ),
  "\n"
)

print(
  unexplained_missing_cone_sites_summer_2025
)


cat(
  "Cone-count sites without a bird survey:",
  nrow(
    cone_sites_without_bird_surveys_summer_2025
  ),
  "\n"
)

print(
  cone_sites_without_bird_surveys_summer_2025
)


if (nrow(unexplained_missing_cone_sites_summer_2025) > 0) {
  stop(
    paste0(
      nrow(unexplained_missing_cone_sites_summer_2025),
      " bird-survey site(s) lack cone data without a documented explanation. ",
      "Review the printed site IDs before continuing."
    )
  )
}


if (nrow(cone_sites_without_bird_surveys_summer_2025) > 0) {
  stop(
    paste0(
      nrow(cone_sites_without_bird_surveys_summer_2025),
      " cone-count site(s) still do not match a bird-survey site. ",
      "Review the printed site IDs before continuing."
    )
  )
}

#### Create Liberal Summer 2025 Presence Data ####
# Liberal presence includes:
# 1. direct CACR detection, or
# 2. crossbill-opened cone evidence.
#
# Sites without a completed cone survey are classified using the strict bird-survey result only.

liberal_summer_2025 <- strict_summer_2025 |>
  left_join(
    cone_evidence_summer_2025,
    by = "Site"
  ) |>
  left_join(
    cone_survey_exceptions_summer_2025,
    by = "Site"
  ) |>
  mutate(
    Cone_Evidence = case_when(
      !is.na(Cone_Evidence) ~
        Cone_Evidence,

      Site %in% c(
        "MAC-058",
        "LPO-060",
        "MAC-089"
      ) ~
        FALSE,

      TRUE ~
        NA
    ),

    Liberal_Presence = as.integer(
      Strict_Presence == 1 |
        Cone_Evidence
    ),

    Presence_Evidence = case_when(
      Strict_Presence == 1 &
        Cone_Evidence ~
        "Detected CACR and cone evidence",

      Strict_Presence == 1 &
        Site %in% c(
          "MAC-058",
          "LPO-060",
          "MAC-089"
        ) ~
        "Detected CACR; classified from bird survey only",

      Strict_Presence == 1 ~
        "Detected CACR only",

      Cone_Evidence ~
        "Cone evidence only",

      Site == "MAC-089" ~
        "No CACR detected; no lodgepole present",

      Site %in% c(
        "MAC-058",
        "LPO-060"
      ) ~
        "No CACR detected; cone survey not completed",

      TRUE ~
        "No presence evidence"
    )
  ) |>
  arrange(
    Site
  )

#### Summarize Summer 2025 Presence Data ####

cat(
  "Total Summer 2025 survey sites:",
  nrow(
    liberal_summer_2025
  ),
  "\n"
)

cat(
  "Strict presences:",
  sum(
    liberal_summer_2025$Strict_Presence == 1,
    na.rm = TRUE
  ),
  "\n"
)

cat(
  "Strict absences:",
  sum(
    liberal_summer_2025$Strict_Presence == 0,
    na.rm = TRUE
  ),
  "\n"
)

cat(
  "Liberal presences:",
  sum(
    liberal_summer_2025$Liberal_Presence == 1,
    na.rm = TRUE
  ),
  "\n"
)

cat(
  "Liberal absences:",
  sum(
    liberal_summer_2025$Liberal_Presence == 0,
    na.rm = TRUE
  ),
  "\n"
)

cat(
  "Liberal classifications unavailable:",
  sum(
    is.na(
      liberal_summer_2025$Liberal_Presence
    )
  ),
  "\n"
)


print(
  liberal_summer_2025 |>
    count(
      Presence_Evidence,
      name = "Sites"
    )
)

print(liberal_summer_2025 |>
  filter(
    Presence_Evidence == "No presence evidence"
  ) |>
  select(
    Site,
    Strict_Presence,
    Cone_Evidence,
    Trees_With_Cone_Counts,
    Trees_With_Cone_Evidence
  ))

#### QA Summer 2025 Survey Corrections and Presence Data ####
# Confirm that all known site corrections and documented exceptions were applied correctly before saving the final strict and liberal datasets.


#### QA 1: Confirm MAC-124 Duplicate Correction ####

mac124_final_check_summer_2025 <- survey_summer_2025 |>
  mutate(
    Site = str_trim(
      str_to_upper(
        as.character(`Survey Point Number`)
      )
    )
  ) |>
  filter(
    Site == "MAC-124"
  ) |>
  select(
    `Date and Time`,
    Site,
    Site_Originally_Recorded,
    Site_ID_Corrected,
    Requires_Manual_Review,
    x,
    y
  )

cat(
  "Final records labeled MAC-124:",
  nrow(mac124_final_check_summer_2025),
  "\n"
)

print(
  mac124_final_check_summer_2025
)

if (nrow(mac124_final_check_summer_2025) != 1) {
  stop(
    paste0(
      "Expected exactly one final survey record labeled MAC-124, but found ",
      nrow(mac124_final_check_summer_2025),
      ". Review the MAC-124 correction."
    )
  )
}

if (any(
  survey_summer_2025$Requires_Manual_Review,
  na.rm = TRUE
)) {
  print(
    survey_summer_2025 |>
      filter(
        Requires_Manual_Review
      ) |>
      select(
        `Date and Time`,
        Site_Originally_Recorded,
        `Survey Point Number`,
        x,
        y
      )
  )

  stop(
    "At least one survey-site correction still requires manual review."
  )
}


#### QA 2: Confirm MAC-079 Was Corrected to MAC-072 ####

cone_survey_ids_summer_2025 <- cone_survey_summer_2025 |>
  transmute(
    Site = str_remove(
      str_trim(
        str_to_upper(
          as.character(`Survey Point ID`)
        )
      ),
      "-D$"
    )
  )

logger_cone_ids_summer_2025 <- logger_cones_summer_2025 |>
  transmute(
    Site = str_remove(
      str_trim(
        str_to_upper(
          as.character(`Survey Point ID`)
        )
      ),
      "-D$"
    )
  )

remaining_mac079_summer_2025 <- bind_rows(
  cone_survey_ids_summer_2025 |>
    mutate(Source = "Cone survey"),

  logger_cone_ids_summer_2025 |>
    mutate(Source = "Logger cone survey")
) |>
  filter(
    Site == "MAC-079"
  )

cat(
  "Remaining cone records labeled MAC-079:",
  nrow(remaining_mac079_summer_2025),
  "\n"
)

print(
  remaining_mac079_summer_2025
)

if (nrow(remaining_mac079_summer_2025) > 0) {
  stop(
    "MAC-079 still appears in the cone data. The correction to MAC-072 was not fully applied."
  )
}

if (!"MAC-072" %in% cone_evidence_summer_2025$Site) {
  stop(
    "MAC-072 does not appear in the final site-level cone-evidence data."
  )
}


#### QA 3: Confirm Documented Cone-Survey Exceptions ####

expected_cone_exceptions_summer_2025 <- tibble(
  Site = c(
    "MAC-058",
    "LPO-060",
    "MAC-089"
  ),

  Expected_Reason = c(
    "Unsafe unauthorized camp activity",
    "Logger could not be located",
    "No lodgepole pine present"
  )
)

missing_exception_records_summer_2025 <-
  expected_cone_exceptions_summer_2025 |>
  anti_join(
    cone_survey_exceptions_summer_2025 |>
      distinct(
        Site
      ),
    by = "Site"
  )

cat(
  "Expected exception sites missing from the exception table:",
  nrow(missing_exception_records_summer_2025),
  "\n"
)

print(
  missing_exception_records_summer_2025
)

if (nrow(missing_exception_records_summer_2025) > 0) {
  stop(
    "One or more documented cone-survey exception sites are missing."
  )
}


unexpected_exception_records_summer_2025 <-
  cone_survey_exceptions_summer_2025 |>
  anti_join(
    expected_cone_exceptions_summer_2025 |>
      distinct(
        Site
      ),
    by = "Site"
  )

cat(
  "Unexpected sites in the cone-survey exception table:",
  nrow(unexpected_exception_records_summer_2025),
  "\n"
)

print(
  unexpected_exception_records_summer_2025
)

if (nrow(unexpected_exception_records_summer_2025) > 0) {
  stop(
    "The cone-survey exception table contains unexpected sites."
  )
}


#### QA 4: Confirm Strict-Only Classification for Exception Sites ####

exception_presence_check_summer_2025 <-
  liberal_summer_2025 |>
  filter(
    Site %in% c(
      "MAC-058",
      "LPO-060",
      "MAC-089"
    )
  ) |>
  select(
    Site,
    Strict_Presence,
    Cone_Evidence,
    Liberal_Presence,
    Presence_Evidence,
    Cone_Survey_Status,
    Cone_Survey_Reason
  )

print(
  exception_presence_check_summer_2025
)

if (nrow(exception_presence_check_summer_2025) != 3) {
  stop(
    paste0(
      "Expected three documented exception sites in the liberal dataset, but found ",
      nrow(exception_presence_check_summer_2025),
      "."
    )
  )
}

incorrect_exception_classifications_summer_2025 <-
  exception_presence_check_summer_2025 |>
  filter(
    is.na(Liberal_Presence) |
      Liberal_Presence != Strict_Presence |
      is.na(Cone_Evidence) |
      Cone_Evidence != FALSE
  )

if (nrow(incorrect_exception_classifications_summer_2025) > 0) {
  print(
    incorrect_exception_classifications_summer_2025
  )

  stop(
    paste0(
      nrow(incorrect_exception_classifications_summer_2025),
      " documented exception site(s) were not classified using the strict survey result only."
    )
  )
}


#### QA 5: Confirm No Unexplained Bird/Cone Mismatches Remain ####

unexplained_bird_sites_without_cones_summer_2025 <-
  strict_summer_2025 |>
  distinct(
    Site
  ) |>
  anti_join(
    cone_evidence_summer_2025 |>
      distinct(
        Site
      ),
    by = "Site"
  ) |>
  anti_join(
    cone_survey_exceptions_summer_2025 |>
      distinct(
        Site
      ),
    by = "Site"
  )

unmatched_cone_sites_summer_2025 <-
  cone_evidence_summer_2025 |>
  distinct(
    Site
  ) |>
  anti_join(
    strict_summer_2025 |>
      distinct(
        Site
      ),
    by = "Site"
  )

cat(
  "Unexplained bird-survey sites without cone data:",
  nrow(unexplained_bird_sites_without_cones_summer_2025),
  "\n"
)

print(
  unexplained_bird_sites_without_cones_summer_2025
)

cat(
  "Cone-data sites without a bird survey:",
  nrow(unmatched_cone_sites_summer_2025),
  "\n"
)

print(
  unmatched_cone_sites_summer_2025
)

if (nrow(unexplained_bird_sites_without_cones_summer_2025) > 0) {
  stop(
    "At least one bird-survey site lacks cone data without a documented exception."
  )
}

if (nrow(unmatched_cone_sites_summer_2025) > 0) {
  stop(
    "At least one cone-data site does not match a bird-survey site."
  )
}


#### QA 6: Confirm Final Presence Values Are Complete ####

invalid_strict_values_summer_2025 <-
  strict_summer_2025 |>
  filter(
    is.na(Strict_Presence) |
      !Strict_Presence %in% c(
        0L,
        1L
      )
  )

invalid_liberal_values_summer_2025 <-
  liberal_summer_2025 |>
  filter(
    is.na(Liberal_Presence) |
      !Liberal_Presence %in% c(
        0L,
        1L
      )
  )

if (nrow(invalid_strict_values_summer_2025) > 0) {
  print(
    invalid_strict_values_summer_2025
  )

  stop(
    "The strict-presence dataset contains missing or invalid values."
  )
}

if (nrow(invalid_liberal_values_summer_2025) > 0) {
  print(
    invalid_liberal_values_summer_2025
  )

  stop(
    "The liberal-presence dataset contains missing or invalid values."
  )
}


#### QA 7: Confirm Liberal Presence Never Removes a Strict Presence ####

strict_liberal_conflicts_summer_2025 <-
  liberal_summer_2025 |>
  filter(
    Strict_Presence == 1 &
      Liberal_Presence != 1
  )

if (nrow(strict_liberal_conflicts_summer_2025) > 0) {
  print(
    strict_liberal_conflicts_summer_2025
  )

  stop(
    "At least one strict presence was not retained as a liberal presence."
  )
}


cat(
  "\nSummer 2025 survey correction and presence QA passed.\n",
  "MAC-124 correction confirmed.\n",
  "MAC-079 to MAC-072 correction confirmed.\n",
  "Documented cone-survey exceptions confirmed.\n",
  "No unexplained bird/cone site mismatches remain.\n",
  "All strict and liberal classifications are complete 0/1 values.\n"
)

#### Save Summer 2025 Presence Data ####

dir.create(
  file.path(
    output_dir,
    "Presence_Data"
  ),
  recursive = TRUE,
  showWarnings = FALSE
)


write_csv(
  strict_summer_2025,
  file.path(
    output_dir,
    "Presence_Data",
    "Strict_Presence_Summer_2025.csv"
  )
)


write_csv(
  liberal_summer_2025,
  file.path(
    output_dir,
    "Presence_Data",
    "Liberal_Presence_Summer_2025.csv"
  )
)

cat(
  "Saved:",
  file.path(
    output_dir,
    "Presence_Data",
    "Strict_Presence_Summer_2025.csv"
  ),
  "\n"
)

cat(
  "Saved:",
  file.path(
    output_dir,
    "Presence_Data",
    "Liberal_Presence_Summer_2025.csv"
  ),
  "\n"
)
```


### Fall
